---
title: "CDS Child Expenditures"
output: html_notebook
---

```{r}
library(dplyr)
library(tidyr)
```

# Child-Specific Expenditures

This script selects the relevant variables on child-specific expenditures from the CDS.

## Variable codes
Vname      Variable Description
---------- ------------------------------------------
KIDS       Sum of All KID Flags                    
KID02      KID2002 = 1 if exists, else missing     
KID07      KID2007 = 1 if exists, else missing     
ER30001    1968 INTERVIEW NUMBER                   
ER30002    PERSON NUMBER                         68
ER33601    2001 INTERVIEW NUMBER                   
ER33602    SEQUENCE NUMBER                       01
ER33603    RELATION TO HEAD                      01
CHREL      PCG CHILD FILE RELEASE NUMBER 02        
Q21B12_2   PRIV SCH PREV YR 02                     
Q21B12A1   SCH COSTS PREV YR - AMT 02              
Q21B12A2   SCH COSTS PREV YR - PER 02              
Q21G5      IN TUTORING PROGRAM 6-9 YRS 02          
Q21G5D     COST OF TUTORING 6-9 YRS 02             
Q21G7A     TAKE LESSONS 6-9 YRS 02                 
Q21G7C     COST OF LESSONS 6-9 YRS 02              
Q21G8C     INVOLVD IN SPORTS LAST SUMMER 6-9 YRS 02
Q21G8F     INVOLVD IN SPORTS LAST SCH YR 6-9 YRS 02
Q21G8I     COST OF SPORTS PAST 12 MTHS 6-9 YRS 02  
Q21G10     MEMBER OF COMMUNITY GRPS 6-9 YRS 02     
Q21G10C    COST OF COMMUNITY GRPS 6-9 YRS 02       
Q21H5      IN TUTORING PROGRAM 10+ YRS 02          
Q21H5C     COST OF TUTORING 10+ YRS 02             
Q21H6      TAKE LESSONS 10+ YRS 02                 
Q21H6B     COST OF LESSONS 10+ YRS 02              
Q21H7B     INVOLVD IN SPORTS LAST SUMMER 10+ YRS 02
Q21H7D     INVOLVD IN SPORTS LAST SCH YR 10+ YRS 02
Q21H7F     COST OF SPORTS PAST 12 MTHS 10+ YRS 02  
Q21H8      MEMBER OF COMMUNITY GRPS 10+ YRS 02     
Q21H8B     COST OF COMMUNITY GRPS 10+ YRS 02       
Q21H23A    SPEND ON TOYS 02                        
Q21H23A1   AMT SPENT ON TOYS 02                    
Q21H23B    SPEND ON VACATION 02                    
Q21H23B1   AMT SPENT OF VACATION 02                
Q21H23C    SPEND ON SCH SUPPLIES 02                
Q21H23C1   AMT SPENT OF SCH SUPPLIES 02            
Q21H23D    SPEND ON CLOTHES 02                     
Q21H23D1   AMT SPENT ON CLOTHES 02                 
Q21H23H    SPEND ON FOOD 02                        
Q21H23H1   AMT SPENT ON FOOD 02                    
ER33901    2007 INTERVIEW NUMBER                   
ER33902    SEQUENCE NUMBER                       07
ER33903    RELATION TO HEAD                      07
PCHREL07   PCG CHILD FILE RELEASE NUMBER 07        
Q31B12_2   PRIV SCH PREV YR 07                     
Q31B12A1   PREV YR SCHOOL COSTS - AMT 07           
Q31B12A2   PREV YR SCHOOL COSTS - PER 07           
Q31H5      IN TUTORING PROG 10+ YRS 07             
Q31H5C     COST OF TUTORING 10+ YRS 07             
Q31H6      TAKE LESSONS 10+ YRS 07                 
Q31H6B     COST OF LESSONS 10+ YRS 07              
Q31H7      INVOLVED SPORTS THIS SCHOOL YR 10+YRS 07
Q31H7D     INVOLVED IN SPORTS LAST SUMMER 10+YRS 07
Q31H7F     COST OF SPORTS PAST 12MOS 10+ YRS 07    
Q31H8      MEMBER OF COMMUNITY GROUPS 10+ YRS 07   
Q31H8B     COST OF COMM GRPS 10+ YRS 07            
Q31H23A    SPEND ON TOYS 07                        
Q31H23A1   AMT SPENT ON TOYS 07                    
Q31H23B    SPEND ON VACATION 07                    
Q31H23B1   AMT SPENT ON VACATION 07                
Q31H23C    SPEND ON SCHOOL SUPPLIES 07             
Q31H23C1   AMT SPENT ON SCHOOL SUPPLIES 07         
Q31H23D    SPEND ON CLOTHES 07                     
Q31H23D1   AMT SPENT ON CLOTHES 07                

These questions can be found in the PCG interview. 

```{r}
#read csv file with the household expenditure information listed above
E <- read.csv("../../../data-cds/expenditures/Exp_0207.csv") %>%
  mutate(KID = ER30001*1000 + ER30002)

#First select kids in 2002 sample and variables describing 2002 hh expenditures

E02 <- E[E$KID02==1.1,] %>%
  select(KID,starts_with("Q21"))

#Substitute to missing if the primary caregiver did not know about costs
E02[(E02$Q21H23A>=8|is.na(E02$Q21H23A)==1|E02$Q21H23A1>=99998),"Q21H23A1"]<-NA
E02[(E02$Q21H23B>=8|is.na(E02$Q21H23B)==1|E02$Q21H23B1>=99998),"Q21H23B1"]<-NA
E02[(E02$Q21H23C>=8|is.na(E02$Q21H23C)==1|E02$Q21H23C1>=9998),"Q21H23C1"]<-NA
E02[(E02$Q21H23D>=8|is.na(E02$Q21H23D)==1|E02$Q21H23D1>=9998),"Q21H23D1"]<-NA
E02[(E02$Q21H23H>=8|is.na(E02$Q21H23H)==1|E02$Q21H23H1>=99998),"Q21H23H1"]<-NA

E02 <- E02 %>%
  rename(Toys = Q21H23A1, Vacation = Q21H23B1, SchSupplies = Q21H23C1, Clothing = Q21H23D1, Food = Q21H23H1) %>%
  mutate(Toys = Toys/52, Vacation = Vacation/52, SchSupplies = SchSupplies/52, Clothing = Clothing/52, Food = Food/52)


#Substitute to missing if the primary caregiver did not know about costs
E02$Q21G5D[(E02$Q21G5D>=9998)] <- NA
E02$Q21H5C[(E02$Q21H5C>=9998)] <- NA
E02$Q21G10C[(E02$Q21G10C>=9998)] <- NA
E02$Q21H8B[(E02$Q21H8B>=9998)] <- NA
E02$Q21G8I[(E02$Q21G8I>=9998)] <- NA
E02$Q21H7F[(E02$Q21H7F>=9998)] <- NA
E02$Q21G7C [(E02$Q21G7C >=9998)] <- NA
E02$Q21H6B[(E02$Q21H6B>=9998)] <- NA
E02$Q21B12A1[(E02$Q21B12A1>=99998)] <- NA


#Combine expenditure on children in diffenet age groups, generate annual tuition cost
E02 <- E02 %>%
  mutate(year = 2002 ) %>%
  mutate(tutoring = Q21G5D + Q21H5C) %>%
  mutate(comm_grps = (Q21G10C + Q21H8B)) %>%
  mutate(sports = (Q21G8I + Q21H7F)) %>%
  mutate(lessons = (Q21G7C + Q21H6B)) %>%
  mutate(tuition = Q21B12A1)%>%
  mutate(tuition = case_when(Q21B12A2==3 ~ 50*as.numeric(tuition),Q21B12A2==5 ~ 12*as.numeric(tuition), TRUE ~ as.numeric(tuition)))%>%
  select(KID, year, tuition, lessons, tutoring, comm_grps, sports, Toys, Vacation, SchSupplies, Clothing, Food)
    

#Select kids in 2007 sample and variables describing 2007 hh expenditures

E07<-E[E$KID07==1.1,] %>%select(KID,starts_with("Q31"))
#Substitute to missing if the primary caregiver did not not about costs
E07[(E07$Q31H23A>=8|is.na(E07$Q31H23A)==1|E07$Q31H23A1>=99998),"Q31H23A1"]<-NA
E07[(E07$Q31H23B>=8|is.na(E07$Q31H23B)==1|E07$Q31H23B1>=99998),"Q31H23B1"]<-NA
E07[(E07$Q31H23C>=8|is.na(E07$Q31H23C)==1|E07$Q31H23C1>=99998),"Q31H23C1"]<-NA
E07[(E07$Q31H23D>=8|is.na(E07$Q31H23D)==1|E07$Q31H23D1>=99998),"Q31H23D1"]<-NA
E07[(E07$Q31H23H>=8|is.na(E07$Q31H23H)==1|E07$Q31H23H1>=99998),"Q31H23H1"]<-NA

E07 <- E07 %>%
  rename(Toys = Q31H23A1, Vacation = Q31H23B1, SchSupplies = Q31H23C1, Clothing = Q31H23D1, Food = Q31H23H1) %>%
  mutate(Toys = Toys/52, Vacation = Vacation/52, SchSupplies = SchSupplies/52, Clothing = Clothing/52, Food = NA)



E07$Q31H5C[(E07$Q31H5C>=9998)] <- NA
E07$Q31H8B[(E07$Q31H8B>=9998)] <- NA
E07$Q31H7F[(E07$Q31H7F>=9998)] <- NA
E07$Q31H6B[(E07$Q31H6B>=9998)] <- NA
E07$Q31B12A1[(E07$Q31B12A1>=99998)] <- NA

#In 2007 children the information is collected for children 10+, generate annual tuition cost
E07 <- E07 %>%
  mutate(year = 2007) %>%
  mutate(tutoring = Q31H5C) %>%
  mutate(comm_grps = Q31H8B) %>%
  mutate(sports = Q31H7F) %>%
  mutate(lessons = Q31H6B) %>%
  mutate(tuition = Q31B12A1)%>%
  mutate(tuition = case_when(Q31B12A2==3 ~ 50*as.numeric(tuition),Q31B12A2==5 ~ 12*as.numeric(tuition), TRUE ~ as.numeric(tuition)))%>%
  select(KID,year,Toys,Vacation,SchSupplies,Clothing,Food,sports,lessons,comm_grps,tutoring,tuition)

#Merge 2002 and 2007 expenditure data, transform annual expenditures to weekly

E <- rbind(E07, E02)
E <- E %>% mutate(tuition=tuition/52, 
         comm_grps=comm_grps/52,
         lessons=lessons/52,
         tutoring=tutoring/52,
         sports=sports/52)

#Drop children with missing ID
E <- E[is.na(E$KID)==0,]

cols <- c("KID","year","Toys","Vacation","SchSupplies","Clothing","Food","sports","lessons","comm_grps","tutoring","tuition")


write.csv(E[,cols],"../../../data-cds/expenditures/Expenditures.csv")
```

# Per Pupil Expenditure for the Child's School

When possible, as part of the CDS supplement, the PSID conducted interviews with the Middle and Elementary School administrators. The data below cleans all available data on reported per pupil expenditure for the school.

```{r}
PPE <- readxl::read_excel("../../../data-cds/expenditures/CDS_PPE.xlsx") %>%
  mutate(KID = ER30001*1000 + ER30002) %>%
  rename(PPE = Q12A27) %>%
  mutate(PPE = case_when(PPE<99998 ~ PPE)) %>% #<- code in missing values
  select(KID,PPE) %>%
  mutate(year = 1997) #<- add the year for merging into a panel.

write.csv(PPE,"../../../data-cds/expenditures/PPE.csv")
```

